\documentclass{article}
\usepackage{graphicx}
\usepackage[margin=2.5cm]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{enumitem}

% Code listing style
\lstset{
    basicstyle=\ttfamily\small,
    backgroundcolor=\color{gray!10},
    frame=single,
    rulecolor=\color{gray!40},
    breaklines=true,
    breakatwhitespace=true,
    showstringspaces=false,
    numbers=none,
    xleftmargin=4pt,
    xrightmargin=4pt,
    framesep=4pt,
    framexleftmargin=4pt
}

\title{Helyx Residual Monitor \\ \large Deployment and User Guide}
\author{Ed Handford}
\date{March 2026}

\begin{document}

\maketitle

\tableofcontents
\newpage

% ============================================================
\section{Introduction}

The \textbf{Helyx Residual Monitor} is a lightweight GUI tool that monitors a running Helyx (or OpenFOAM) solver in real time. It watches the solver log file for velocity residual values and automatically stops the simulation when all monitored residuals drop below a user-defined convergence threshold.

This removes the need to manually watch residual plots or set overly conservative iteration counts --- the solver runs for exactly as long as it needs to converge, then stops cleanly.

\subsection{Key Features}

\begin{itemize}[nosep]
    \item Small, standalone GUI --- no complex setup required.
    \item Monitors velocity residuals (\texttt{Ux}, \texttt{Uy}, \texttt{Uz}) in real time.
    \item User-configurable convergence threshold (e.g.\ \texttt{1e-4}).
    \item Automatically patches \texttt{system/controlDict} to stop the solver cleanly.
    \item Creates a timestamped backup of \texttt{controlDict} before any modification.
    \item Zero third-party dependencies --- uses only the Python standard library.
\end{itemize}

% ============================================================
\section{Prerequisites}

Before deploying the tool, ensure the following are available on the target machine.

\subsection{For Running from Source}

\begin{itemize}[nosep]
    \item \textbf{Python 3.9 or later} --- download from \url{https://www.python.org/downloads/}.
    \item During installation on Windows, tick \textbf{``Add Python to PATH''} and ensure the \textbf{tkinter} component is selected (it is included by default).
    \item No additional packages (\texttt{pip install}) are required.
\end{itemize}

\subsection{For Building a Standalone Executable}

\begin{itemize}[nosep]
    \item Everything listed above, plus:
    \item \textbf{PyInstaller} --- installed via \texttt{pip install pyinstaller}.
\end{itemize}

% ============================================================
\section{Deployment}

There are two ways to deploy the tool: running directly from the Python source file, or building a standalone Windows executable.

\subsection{Option A: Run from Source (Recommended for Development)}

\begin{enumerate}[nosep]
    \item Copy \texttt{helyx\_monitor.py} to any convenient location on your machine.
    \item Open a terminal (Command Prompt or PowerShell) and navigate to that location.
    \item Run:
\begin{lstlisting}
python helyx_monitor.py
\end{lstlisting}
    \item The GUI window will appear and is ready to use.
\end{enumerate}

\subsection{Option B: Build a Standalone Executable (Recommended for Distribution)}

This produces a single \texttt{.exe} file that can be shared with colleagues who may not have Python installed.

\begin{enumerate}[nosep]
    \item Open a terminal in the project directory (where \texttt{helyx\_monitor.py} is located).
    \item Install PyInstaller if not already installed:
\begin{lstlisting}
pip install pyinstaller
\end{lstlisting}
    \item Run the provided build script:
\begin{lstlisting}
build_windows.bat
\end{lstlisting}
    Alternatively, run the PyInstaller command directly:
\begin{lstlisting}
pyinstaller --onefile --windowed --name "HelyxResidualMonitor" helyx_monitor.py
\end{lstlisting}
    \item The executable will be created at:
\begin{lstlisting}
dist\HelyxResidualMonitor.exe
\end{lstlisting}
    \item Copy this single file to any machine. No installation is needed --- double-click to run.
\end{enumerate}

\noindent The resulting executable is approximately 10\,MB and includes an embedded Python runtime.

% ============================================================
\section{Usage}

\subsection{Step-by-Step Workflow}

\begin{enumerate}
    \item \textbf{Launch the tool.} \\
    Run \texttt{helyx\_monitor.py} (or double-click the \texttt{.exe}).

    \item \textbf{Select the case directory.} \\
    Click \emph{Browse} next to ``Case Directory'' and navigate to the root of your Helyx/OpenFOAM case (the folder that contains the \texttt{system/}, \texttt{constant/}, and time directories).

    \item \textbf{Verify the log file.} \\
    The tool auto-detects the most recent log file in the case directory (e.g.\ \texttt{log.simpleFoam}). If this is incorrect, click \emph{Browse} next to ``Log File'' to select the correct one manually.

    \item \textbf{Set the convergence threshold.} \\
    Enter the desired residual threshold in the ``Threshold'' field. Common values:
    \begin{itemize}[nosep]
        \item \texttt{1e-4} --- loosely converged
        \item \texttt{1e-5} --- well converged
        \item \texttt{1e-6} --- tightly converged
    \end{itemize}

    \item \textbf{Select variables to monitor.} \\
    Tick or untick the velocity components (\texttt{Ux}, \texttt{Uy}, \texttt{Uz}). All three are selected by default. The solver will only be stopped when \emph{all} selected variables are below the threshold simultaneously.

    \item \textbf{Start the solver in Helyx.} \\
    Use the Helyx GUI to begin your simulation as normal.

    \item \textbf{Click ``Start Monitoring''.} \\
    The tool begins tailing the log file. Current residual values are displayed and colour-coded:
    \begin{itemize}[nosep]
        \item \textcolor{red}{Red} --- above the threshold.
        \item \textcolor{green}{Green} --- below the threshold.
    \end{itemize}

    \item \textbf{Wait for convergence.} \\
    When all monitored residuals drop below the threshold, the tool will:
    \begin{enumerate}[nosep]
        \item Create a backup of \texttt{system/controlDict} (named \texttt{controlDict.bak\_YYYYMMDD\_HHMMSS}).
        \item Patch \texttt{controlDict} to set \texttt{stopAt writeNow;}.
        \item Display a notification dialog confirming the solver has been stopped.
    \end{enumerate}
    The Helyx solver will finish its current timestep, write the results, and exit cleanly.
\end{enumerate}

\subsection{Stopping Manually}

Click \emph{Stop Monitoring} at any time to stop the tool without affecting the solver. The solver will continue running --- only the monitoring is halted.

% ============================================================
\section{How It Works}

This section describes the tool's internal operation for reference.

\subsection{Log File Parsing}

The tool reads new data from the solver log file every 500\,ms. It matches lines against the standard OpenFOAM residual format:

\begin{lstlisting}
smoothSolver:  Solving for Ux, Initial residual = 0.0023, Final residual = 1.2e-05, No Iterations 3
\end{lstlisting}

The \textbf{Initial residual} value is extracted. This is the standard convergence metric in CFD --- it represents the residual at the start of each solver iteration, before the linear system is solved. The Final residual (post-solve) is not used, as it can give a misleading impression of convergence.

\subsection{Solver Stop Mechanism}

When the convergence threshold is met, the tool modifies the \texttt{stopAt} entry in \texttt{system/controlDict} to \texttt{writeNow}. OpenFOAM-based solvers (including Helyx) re-read \texttt{controlDict} between timesteps, so this causes the solver to:

\begin{enumerate}[nosep]
    \item Complete the current timestep.
    \item Write the solution data to disk.
    \item Exit cleanly.
\end{enumerate}

The modification is performed atomically (write to a temporary file, then rename) to avoid corrupting \texttt{controlDict} if the solver reads it mid-write.

\subsection{Backup and Recovery}

Before any modification, a timestamped copy of the original \texttt{controlDict} is saved (e.g.\ \texttt{controlDict.bak\_20260315\_143022}). To restore the original after a run, simply rename the backup file back to \texttt{controlDict}, or change \texttt{stopAt} back to \texttt{endTime} manually.

% ============================================================
\section{Troubleshooting}

\begin{table}[h]
\centering
\begin{tabular}{@{}p{5.5cm}p{8.5cm}@{}}
\toprule
\textbf{Symptom} & \textbf{Solution} \\
\midrule
``Warning: no residuals detected'' appears after 30 seconds &
The log file may not contain solver output yet. Ensure the Helyx solver is running and that the correct log file is selected. \\
\addlinespace
``system/controlDict not found'' error on start &
The selected directory is not a valid OpenFOAM case. Browse to the root case folder (the one containing \texttt{system/}, \texttt{constant/}, etc.). \\
\addlinespace
``Could not write to controlDict'' error &
The file may be read-only or locked. Check file permissions in Windows Explorer (right-click $\to$ Properties $\to$ uncheck ``Read-only''). \\
\addlinespace
Solver does not stop after threshold is reached &
Verify that \texttt{system/controlDict} now contains \texttt{stopAt writeNow;}. Some custom solver configurations may not re-read \texttt{controlDict} during a run. \\
\addlinespace
GUI does not open (no window appears) &
Ensure Python was installed with tkinter. Run \texttt{python -c "import tkinter"} in a terminal --- if this produces an error, reinstall Python with the tkinter option selected. \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{File Reference}

\begin{table}[h]
\centering
\begin{tabular}{@{}lp{9cm}@{}}
\toprule
\textbf{File} & \textbf{Description} \\
\midrule
\texttt{helyx\_monitor.py} & The complete application (GUI and backend logic in a single file). \\
\texttt{build\_windows.bat} & One-click build script that invokes PyInstaller to produce \texttt{HelyxResidualMonitor.exe}. \\
\texttt{main.tex} & This documentation (LaTeX source for Overleaf). \\
\texttt{README.md} & Quick-start guide. \\
\texttt{.gitignore} & Excludes Python cache and build artifacts from version control. \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
